Reducing Horton-Strahler Stream Order Can Enhance Flood Inundation Mapping Skill with Applications for the U.S. National Water Model

### 2022-01-10 manuscripts/owp_fim4_2021/sections/materials_and_methods.tex
edits after odgen/salas comments

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
OWP FIM `Cahaba' is a fully operational pipeline of software tools to help acquire datasets, cache hydrofabrics, produce FIMs, and evaluate results.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Software Dependencies and Architecture}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
OWP FIM `Cahaba' exclusively utilizes free and open source software dependencies including Python 3, GDAL, TauDEM, Geographic Resource Analysis Support System (GRASS), GNU Parallel, and MPICH \cite{python382,gdal2020,tarboton2005terrain,grass2020,tange2015gnu,amer2021mpich}.
Within the Python 3 ecosystem, many common packages are employed including but not limited to RichDEM, GeoPandas, Rasterio, Rasterstats, and Numba \cite{barnes2018richdem,jordahl2014geopandas,lam2015numba}. 
To simplify setup and enhance portability across host operating systems OWP FIM packages all dependencies up in a Docker image (\url{https://docs.docker.com/engine/install/}). 
A user only need to install Docker on their host machine and build the image from the provided recipe. 
Source code is made available for this project on GitHub (https://github.com/NOAA-OWP/cahaba).

@@

@@

Instructions for accessing data are provided on the project's GitHub page via an Amazon Web Services (AWS) S3 bucket.

@@

For enforcing levee data into the NED DEMs, the USACE National Levee Database (NLD) is used via burning feature elevations \cite{engineers2016national}.

@@

Additionally, the Base Level Engineering (BLE) datasets spanning parts of 9 states including Colorado, New Mexico, Texas, Oklahoma, Kansas, Arkansas, Louisiana, Missouri and Mississippi at two recurrence intervals, 1\% (100 year) and 0.2\% (500 year), are used as validation in this study and furnished by Region 6 of FEMA \cite{fema2021base,fema2021estimated}. 

@@

USGS & NHDPlusHR DEM & DEM used for NHDPlus HR at 1/3 arc-second (10m) spatial resolution and vertical units in centimeters \\

@@

NOAA-OWP & Waterbodies considered by the NWM as reservoirs or lakes. \\

@@

FEMA & Flood Inundation Extents & Inundation depths produced by FEMA BLE HEC-RAS 1D for 1\% and 0.2\% recurrence interval events. \\

@@

Areas with all the required data (from the NWM and the USGS) are labeled as the FIM domain which includes 2,188 HUC8s for the FR and Generalized Mainstems (GMS) networks and 1,604 HUC8s for the MS method. 

@@

An enhancement of OWP FIM `Cahaba' over previous HAND based FIM versions is the support for Hawaii and Puerto Rico which the NWM v2.1 will cover.

@@

The DEM from the NED is subject to a series of hydro-conditioning procedures to enhance its suitability for riverine flood inundation mapping. 

@@

It can also be used to simulate more accurate bathymetry which is not accounted for in the NED 10m DEM \cite{gesch2002national}.

Specifically within the context of `Cahaba', the hydro-conditioning operations that take place in sequential order are presented. 
Prior to any hydro-conditioning, all input datasets must be subset from their original spatial domain scales into the processing units of size HUC8. 
The subsetting is done by spatial query for the cases of the levees, DEM, and NWM hydrofabric while the NHDPlusHR BurnLineEvents are subset via attribute query for the given reach code's membership in the processing unit.
Hydro-conditioning raster operations take place on buffered boundary definitions to avoid edge contamination and effects \cite{lindsay2013measuring}. 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Stream Network Enforcement} 
\label{ssec:stream_network_enforcment}
%
The location of the stream network is enforced to ensure general agreement with established stream networks.
The NHDPlus HR Beta Burnline Event layer is used to enforce stream locations in the NHDPlusHR workflow so it is also used here for hydro-enforcement \cite{moore2019user}. 
However, to better match the drainage density of the NWM V2.1 stream network which is based on the NHDPlus Medium Resolution, the Burnline Events are pruned utilizing a nearest neighbor search around the NWM flowlines.
For every NWM headwater segment a headwater point is derived and linearly interpolated to the nearest Burnline Event segment.
Burnline Event headwater segments are split at the adjusted headwater point to match NWM flowlines.
The resulting pruned NHD stream network is what gets hydro-enforced in subsequent operations.
This procedure is best illustrated in Figure \ref{fig:stream_density_pruning} which shows that the pruned NHD network corresponds to the full density NHD network at NWM V2.1 headwater locations only. 
Additionally, the NHDPlusHR pruned headwaters are later used for seeding a new FIM drainage network that best agrees with the DEM after all hydro-conditioning takes place.
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/headwaters.jpg}
\caption{Pruning of NHDPlus HR Beta Burnline Events (dotted black) to NWM V2.1 stream density (blue) using nearest neighbor selection and linear interpolation. Resulting stream network (red) matches the drainage density of NWM V2.1 while corresponding spatially with the NHDPlusHR Burnline Events.}
\label{fig:stream_density_pruning}
\end{figure}
%
This results in a stream network that has the same density as the NWM V2.1 flowline network but utilizes the locations of the NHDPlusHR Beta BurnLineEvents.

@@

The National Elevation Dataset at 10m resolution lacks sufficient representation of fine grain features such as embankments, floodwalls, and closure structures.
In order to better represent the influences of these features upon hydraulics and inundation extents, the National Levee Database (NLD) published by USACE was used to enforce elevations within the 10m DEM.

@@

Each zone is defined as the thalweg's pixel nearest neighborhood within a maximum distance of 50m.

@@

Based on previous research, splitting each remaining stream link into equidistant reaches not to exceed a parameterized value of 1.5km helps improve synthetic rating curve and mapping skill \cite{garousi2019terrain,godbout2019error,zheng2018geoflood}.
Small reaches can lead to unrealistic variances in channel geometries while oversized reaches can lead to grouping too much geometry variance into one parameterized unit.

@@

Gages are set at the pixel center points of the delineated stream lines explained in Section \ref{sssec:deriving_fim_stream_network}.
The gages act as root nodes in a tree structure and the connectivity is traversed to derive the contributing area for each gage.

@@

\textbf{Q}[y] = \frac{1}{n} \frac{V[y]^{5/3}S^{1/2}}{L B[y]^{2/3}} 

@@

@@

For two subsequent methods discussed in Sections \ref{sssec:replace_and_route_mainstems} and \ref{sssec:generalized_mainstems}, we find the mid-point of the derived stream reach line described in Section \ref{sssec:deriving_fim_stream_network} and find the NWM catchment that contains the mid-point.
Additionally, only relevant catchments from the NWM for the given level path are selected for cross-walking for methods in Sections \ref{sssec:replace_and_route_mainstems} and \ref{sssec:generalized_mainstems}.

@@

A `Cahaba' FIM is generated using the NWM full-resolution stream network and compared with a FEMA 100 year extent (please see Section \ref{ssec:evaluation} for more details) showing significant under-prediction in inundation extent.
The higher discharge along the mainstem of 1,900 cubic meters per second (CMS) does not translate to the lower flow rates along the tributaries of 84 and 195 CMS. 

@@

As a parallel problem, there is excess water accumulated along the mainstem that cannot extend in either a fluvial or pluvial manner beyond the boundaries of the mainstem catchments.

@@

\caption{False negatives associated with confluence of tributaries with mainstems. Integers represent flow values from BLE 100 year event for the associated areas. 

@@

\subsubsection{NWS Replace and Route Mainstems}
\label{sssec:replace_and_route_mainstems}

@@

The RnR configuration of the NWM seeks to replace NWM forecasts by assimilating best available AHPS forecasts and routing them downstream.
The RnR Mainstems (MS) network is a subset of the NWM full-resolution (FR) network at and downstream of AHPS forecast points as seen in Figure \ref{fig:forecast_points}.
The RnR MS network, or simply MS, comprises about 200 thousand km of stream length which is less than 4\% of the FR total stream length of 5.5 million km.

@@

Within each HUC, you'll typically only find a MS stream network of stream order 1 but this can vary if more than one AHPS forecasting point is found within or upstream of the HUC in question.

@@

To further the efforts implemented by MS, we sought to derive HAND at a level path scale which we call Generalized Mainstems (GMS).

@@

To illustrate the GMS procedure, we reference Figure \ref{fig:gms_methods} to show how deriving HAND and FIMs from GMS works.
In Figure \ref{fig:gms_methods}a, we uniquely color code the level paths derived for the NWM stream network. 
For each one of these lines, we derive HAND and its associated datasets including catchments, crosswalks, and rating curves.
Each level path is buffered to a polygon with a user-available distance parameter of 7km and this polygon is used to subset the original DEM for two selected level paths in Figure \ref{fig:gms_methods}b.
We illustrate two HAND grids for two of the level paths in this HUC8 in Figure \ref{fig:gms_methods}c.
Once the FIM hydrofabrics for each level path, we can inundate them individually also shown in Figure \ref{fig:gms_methods}d.
Lastly these individual FIMs are mosaiced together as explained in Section \ref{ssec:inundation_mapping} and shown in Figure \ref{fig:gms_methods}c.
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/level_path_methods.jpg}
\caption{Illustrates how level paths for the NWM are derived.
Level paths symbolized by lines of unique colors are propagated upstream following the direction of maximum arbolate sum at each junction.
}
\label{fig:level_path_methods}
\end{figure}
%
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/gms_methods.jpg}
\caption{Overall procedure for GMS HAND.
In (a), we illustrate all NWM stream lines symbolized by their level path.
Meanwhile (b), demonstrates the NED DEM clipped to a 7km buffer around two selected level paths.
In (c), we show how HAND can be computed just for each one of these two level paths independently. 
We also show inundation maps created for these two level paths in (c). 
In (d), we show all the inundation maps for all the level paths mosaiced together. }
\label{fig:gms_methods}
\end{figure}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Inundation Mapping}
\label{ssec:inundation_mapping}
%
The FIM hydrofabric consisting of the relative elevations grid, catchments grid, catchment polygons, rating curve, and cross-walking data are all used to convert forecasts from the NWM into forecasts extents.
For operational situations, one would cache the FIM hydrofabric then either produce libraries of FIM for a sample of discharges or stages or also produce the FIM in near real-time (NRT).
From the cached FIM hydrofabric and design or forecast discharges including those extracted from the NWM, inundation maps can be generated at HUC 8 spatial processing units in a rapid, parallel operation. 
The discharges are associated with NWM reach identifiers and cross-walked over to reach identifiers in the FIM hydrofabric.

@@

Equation \ref{eq:comp_fim} illustrates the same process but for the mosaic the binary inundation.

@@

Evaluation of our relative elevation CFIM method is conducted by comparison to the HEC-RAS 1D derived models produced by FEMA region 6 \cite{fema2021base,fema2021estimated}.

@@

The maps to the 1\% recurrence flow (1 in 100 year) and the 0.2\% recurrence flow (1 in 500 year) are furnished by FEMA so we used those corresponding discharges and mapping extents for evaluation.

@@

By using the same HEC-RAS derived discharges and FIM extents, we are able to separate out errors introduced by hydrology, forcings, hydraulic routing, etc that we would have potentially seen if we used NWM forecasted discharges.

@@

We elected to spatially intersect the HEC-RAS cross sections with the NWM stream network assigning the 1\% and 0.2\% flow rates with each NWM reach. 

@@

Additionally, the stream network of the FEMA furnished models are of higher stream densities and bifurcation ratios, as evident in Figure \ref{fig:ble_evaluation_method}, leading to a significant amount of false negatives (FN) (under-prediction) along headwater streams with Horton-Strahler orders of one due to the lack of representation of these additional headwater streams in the NWM network.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
OWP FIM `Cahaba' is a fully operational pipeline of software tools to help acquire datasets, cache hydrofabrics, produce FIMs, and evaluate results.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Software Dependencies and Architecture}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
OWP FIM `Cahaba' exclusively utilizes free and open source software dependencies including Python 3, GDAL, TauDEM, Geographic Resource Analysis Support System (GRASS), GNU Parallel, and MPICH \cite{python382,gdal2020,tarboton2005terrain,grass2020,tange2015gnu,amer2021mpich}.
Within the Python 3 ecosystem, many common packages are employed including but not limited to RichDEM, GeoPandas, Rasterio, Rasterstats, and Numba \cite{barnes2018richdem,jordahl2014geopandas,lam2015numba}. 
To simplify setup and enhance portability across host operating systems OWP FIM packages all dependencies up in a Docker image (\url{https://docs.docker.com/engine/install/}). 
A user only needs to install Docker on their host machine and build the image from the provided recipe. 
Source code is made available for this project on GitHub (https://github.com/NOAA-OWP/cahaba).

@@

Functionality is included to turn FIM hydrofabric and streamflows into inundation maps represented in both binary and depths.
Lastly, a test suite includes means of calculating evaluation metrics compared to a variety of preprocessed test case data.

@@

Instructions for accessing data are provided on the project's GitHub page via an Amazon Web Services (AWS) S3 bucket furnished by Earth Science Information Partners (ESIP).

@@

For enforcing levee data into the DEMs, the USACE National Levee Database (NLD) is used to burn feature elevations \cite{engineers2016national}.

@@

Additionally, the Base Level Engineering (BLE) datasets within FEMA Region 6 spanning parts of 9 states including Colorado, New Mexico, Texas, Oklahoma, Kansas, Arkansas, Louisiana, Missouri and Mississippi at two recurrence intervals, 1\% (100 year) and 0.2\% (500 year), are used as validation in this study and furnished by the Interagency Flood Risk Management (InFRM) consortium \cite{fema2021base,fema2021estimated}. 

@@

USGS & NHDPlusHR DEM & DEM used for NHDPlus HR at 1/3 arc-second (10 m) spatial resolution and vertical units in centimeters \\

@@

NOAA-OWP & NWM Waterbodies & Waterbodies considered by the NWM as reservoirs or lakes. \\

@@

FEMA & Flood Inundation Extents & Inundation depths produced by InFRM BLE HEC-RAS 1D for 1\% and 0.2\% recurrence interval events. \\

@@

Areas with all the required data (from the NWM and the USGS) are labeled as the FIM domain which includes 2,188 HUC8s for the FR and GMS networks and 1,604 HUC8s for the MS method. 

@@

An enhancement of OWP FIM `Cahaba' over previous HAND based FIM versions is the support for Hawaii and Puerto Rico which the NWM V2.1 will cover.

@@

The DEM is subject to a series of hydro-conditioning procedures to enhance its suitability for riverine flood inundation mapping. 

@@

It can also be used to simulate more accurate bathymetry which is not accounted for in the 10 m DEM \cite{gesch2002national}.

Specifically within the context of `Cahaba', the hydro-conditioning operations that take place in sequential order are presented. 
Prior to any hydro-conditioning, all input datasets must be subset from their original spatial domain scales into the processing units of size HUC8. 
The subsetting is done by spatial query for the cases of the levees, DEM, and NWM hydrofabric while the NHDPlusHR BurnLineEvents are subset via attribute query for the given reach code's membership in the processing unit.
Hydro-conditioning raster operations take place on buffered boundary definitions to avoid edge contamination and effects \cite{lindsay2013measuring}. 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Stream Network Enforcement} 
\label{ssec:stream_network_enforcment}
%
The location of the stream network is enforced to ensure general agreement with established stream networks.
The NHDPlusHR Beta Burnline Event layer is used to enforce stream locations in the NHDPlusHR workflow so it is also used here for hydro-enforcement \cite{moore2019user}. 
However, to better match the drainage density of the NWM V2.1 stream network which is based on the NHDPlus Medium Resolution, the Burnline Events are pruned utilizing a nearest neighbor search around the NWM flowlines.
For every NWM headwater segment a headwater point is derived and linearly interpolated to the nearest Burnline Event segment.
Burnline Event headwater segments are split at the adjusted headwater point to match NWM flowlines.
The resulting pruned NHD stream network is what gets hydro-enforced in subsequent operations.
This procedure is best illustrated in Figure \ref{fig:stream_density_pruning} which shows that the pruned NHD network corresponds to the full density NHD network at NWM V2.1 headwater locations only. 
Additionally, the NHDPlusHR pruned headwaters are later used for seeding a new FIM drainage network that best agrees with the DEM after all hydro-conditioning takes place.
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/headwaters.jpg}
\caption{Pruning of NHDPlus HR Beta Burnline Events (dotted black) to NWM V2.1 stream density (blue) using nearest neighbor selection and linear interpolation. Resulting stream network (red) matches the drainage density of NWM V2.1 while corresponding spatially with the NHDPlusHR Burnline Events.}
\label{fig:stream_density_pruning}
\end{figure}
%
This results in a stream network that has the same density as the NWM V2.1 flowline network but utilizes the locations of the NHDPlusHR Beta BurnLineEvents.

@@

The DEM at 10 m resolution lacks sufficient representation of fine grain features such as embankments, floodwalls, and closure structures.
In order to better represent the influences of these features upon hydraulics and inundation extents, the National Levee Database (NLD) published by USACE was used to enforce elevations within the 10 m DEM.

@@

Each zone is defined as the thalweg's pixel nearest neighborhood within a maximum distance of 50 m.

@@

Based on previous research, splitting each remaining stream link into equidistant reaches not to exceed a parameterized value of 1.5 km helps improve synthetic rating curve and mapping skill \cite{garousi2019terrain,godbout2019error,zheng2018geoflood}.
Small reaches can lead to unrealistic variances in channel geometries while oversized reaches can lead to grouping too much slope variance into one discretation of the stream network.

@@

Outlet points are set at the pixel center points of the delineated stream lines explained in Section \ref{sssec:deriving_fim_stream_network}.
The outlets act as root nodes in a tree structure and the connectivity is traversed to derive the contributing area for each gage.

@@

\textbf{Q} = \frac{1}{n} \frac{V^{5/3}S^{1/2}}{L B^{2/3}} 

@@

Q, V, and B are taken a specific y values so are more formally written as $Q = Q(y)$, $V = V(y)$, and $B = B(y)$, respectively.

@@

For two subsequent methods discussed in Sections \ref{sssec:nws_mainstems} and \ref{sssec:generalized_mainstems}, we find the mid-point of the derived stream reach line described in Section \ref{sssec:deriving_fim_stream_network} and find the NWM catchment that contains the mid-point.
Additionally, only relevant catchments from the NWM for the given level path are selected for cross-walking for methods in Sections \ref{sssec:nws_mainstems} and \ref{sssec:generalized_mainstems}.

@@

A `Cahaba' FIM is generated using the NWM full-resolution stream network and compared with a FEMA 100 year extent (see Section \ref{ssec:evaluation} for more details) showing significant under-prediction in inundation extent.
The higher discharge along the MS of 1,900 cubic meters per second (CMS) does not translate to the lower flow rates along the tributaries of 84 and 195 CMS. 

@@

As a parallel problem, there is excess water accumulated along the MS that cannot extend in either a fluvial or pluvial manner beyond the boundaries of the MS catchments.

@@

\caption{False negatives associated with confluence of tributaries with MS. Integers represent flow values from BLE 100 year event for the associated areas. 

@@

\subsubsection{NWS Main-stems}
\label{sssec:nws_mainstems}

@@

The Mainstems (MS) network is a subset of the NWM full-resolution (FR) network at and downstream of AHPS forecast points as seen in Figure \ref{fig:forecast_points}.
The MS network comprises about 200 thousand km of stream length which is less than 4\% of the FR total stream length of 5.5 million km.

@@

Within each HUC, you'll typically only find a MS stream network of stream order 1 (i.e. headwater) but this can vary if more than one AHPS forecasting point is found within or upstream of the HUC in question.

@@

To further the efforts implemented by MS, we sought to derive HAND at a level path scale which we call GMS.

@@

To illustrate the GMS procedure, we reference Figure \ref{fig:gms_methods} to show how deriving HAND and FIMs from GMS works.
In Figure \ref{fig:gms_methods}a, we uniquely color code the level paths derived for the NWM stream network. 
For each one of these lines, we derive HAND and its associated datasets including catchments, crosswalks, and rating curves.
Each level path is buffered to a polygon with a user-available distance parameter of 7 km and this polygon is used to subset the original DEM for two selected level paths in Figure \ref{fig:gms_methods}b.
We illustrate two HAND grids for two of the level paths in this HUC8 in Figure \ref{fig:gms_methods}c.
Once the FIM hydrofabrics for each level path are generated, we can inundate them individually also shown in Figure \ref{fig:gms_methods}d.
Lastly these individual FIMs are mosaiced together as explained in Section \ref{ssec:inundation_mapping} and shown in Figure \ref{fig:gms_methods}c.
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/level_path_methods.jpg}
\caption{Illustrates how level paths for the NWM are derived.
Level paths symbolized by lines of unique colors are propagated upstream following the direction of maximum arbolate sum at each junction.
}
\label{fig:level_path_methods}
\end{figure}
%
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/gms_methods.jpg}
\caption{Overall procedure for GMS HAND.
In (a), we illustrate all NWM stream lines symbolized by their level path.
Meanwhile (b), demonstrates the DEM clipped to a 7 km buffer around two selected level paths.
In (c), we show how HAND can be computed just for each one of these two level paths independently. 
We also show inundation maps created for these two level paths in (c). 
In (d), we show all the inundation maps for all the level paths mosaiced together. }
\label{fig:gms_methods}
\end{figure}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Inundation Mapping}
\label{ssec:inundation_mapping}
%
The FIM hydrofabric consisting of the relative elevations grid, catchments grid, catchment polygons, rating curve, and cross-walking data are all used to convert forecasts from the NWM into forecasts extents.
For operational situations, one would cache the FIM hydrofabric then either produce libraries of FIM for a sample of discharges or stages or also produce the FIM in near real-time (NRT).
From the cached FIM hydrofabric and design or forecast discharges including those extracted from the NWM, inundation maps can be generated at HUC 8 spatial processing units in a rapid, parallel operation. 
The discharges are associated with NWM reach identifiers and cross-walked over to reach identifiers in the FIM hydrofabric.

@@

Equation \ref{eq:comp_fim} illustrates the same process but for mosaicing the binary inundation maps.

@@

Evaluation of our relative elevation CFIM method is conducted by comparison to the HEC-RAS 1D derived models produced within FEMA region 6 \cite{fema2021base,fema2021estimated}.

@@

The maps to the 1\% recurrence flow (1 in 100 year) and the 0.2\% recurrence flow (1 in 500 year) are furnished by InFRM so we used those corresponding discharges and mapping extents for evaluation.

@@

By using the same HEC-RAS derived discharges and FIM extents, we are able to separate out errors introduced by hydrology, atmospheric forcings, hydraulic routing, etc that we would have potentially seen if we used NWM forecasted discharges.

@@

We elected to spatially intersect the HEC-RAS cross sections with the NWM stream network assigning the 1\% and 0.2\% flow rates to each NWM reach. 

@@

Additionally, the stream network of the InFRM furnished models are of higher stream densities and bifurcation ratios, as evident in Figure \ref{fig:ble_evaluation_method}, leading to a significant amount of false negatives (FN) (under-prediction) along headwater streams with Horton-Strahler orders of one due to the lack of representation of these additional headwater streams in the NWM network.

### 2022-01-10 manuscripts/owp_fim4_2021/sections/results.tex
edits after odgen/salas comments

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mapping Performance}
\label{ssec:mapping_performance}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We produced FIMs for the entire BLE domain within the 49 HUC8 area across several states in the south central US. 
The forecasted FIMs using the discharges for the 1\% (100 year) and 0.2\% (500 year) recurrence flows directly from HEC-RAS were used to avoid noise and errors from hydrological processes.
We computed the statistics (CSI, POD, and FAR) for both 100 and 500 year events for Mannings N set to 0.06 and 0.12. 
The distribution of these statistics can be examined in Figure \ref{fig:violin_plot} as violin plots.
Each half of a violin plot represents the kernel density estimation (KDE) for a given model (FR, MS, GMS), given Manning's n value (0.06, 0.12), and given recurrence interval (1\%, 0.2\%), and performance metric (CSI, POD, FAR).
We also denote trend lines for each metric and Manning's n setting as well as their respective slope estimate and one-tailed p-value denoting the level of significance of the trend.

Aggregating the metrics in the method above treats each HUC8 as it's own unit and does little to consider the size differences of the HUCs. 
In an opposing aggregation method, we illustrate in Table \ref{tab:aggregate_metrics} the sum of all the TPs, FPs, and FNs to recompute CSI, POD, and FAR across the entire domain. 
%
\begin{figure}[h!]
\centering
\includegraphics[scale=0.9]{figures/violin_plots.jpg}
\caption{Shows kernel density estimation of the distributions (sample size = 49) for 1\% (100 year) and 0.2\% (500 year) along with horizontal, dashed lines for the 25th, 50th, and 75th percentiles (in order from bottom to top).
The sub-figures separate the combination of three metrics (CSI, POD, and FAR) for two settings of Manning's n (0.06 and 0.12).
Trend lines for each metric - Mannings combination are shown (sample size = 294) along with associated slope and p-value of slope testing one-tailed significance.}
\label{fig:violin_plot}
\end{figure}
%
\begin{table}[h!]
\caption{Primary metrics, TPs, FPs, and FNs, aggregated for BLE domain to recompute CSI, POD, and FAR.}
\label{tab:aggregate_metrics}
\centering
%\begin{tabular}{|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
\begin{tabular}{|c|c||c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Metric} & \multirow{2}{*}{Manning's n} & \multicolumn{2}{|c|}{FR} & \multicolumn{2}{|c|}{MS} & \multicolumn{2}{|c|}{GMS} \\
\cline{3-8}
  &  & 100yr & 500yr & 100yr & 500yr & 100yr & 500yr \\
\hline
\multirow{2}{*}{CSI} & 0.06 & 0.5576 & 0.5839 & 0.5717 & 0.5990 & 0.5796 & 0.6075 \\
\cline{2-8}
  & 0.12 & 0.5915 & 0.6149 & 0.6054 & 0.6288 & 0.6182 & 0.6435 \\
\hline
\multirow{2}{*}{POD} & 0.06 & 0.6354 & 0.6575 & 0.6524 & 0.6755 & 0.6633 & 0.6863 \\
\cline{2-8}
  & 0.12 & 0.7255 & 0.7446 & 0.7460 & 0.7648 & 0.7606 & 0.7810 \\
\hline
\multirow{2}{*}{FAR} & 0.06 & 0.1800 & 0.1609 & 0.1778 & 0.1589 & 0.1787 & 0.1589 \\
\cline{2-8}
  & 0.12 & 0.2379 & 0.2208 & 0.2374 & 0.2204 & 0.2324 & 0.2148 \\
\hline
\end{tabular}
\end{table}
%
% Interpretation of metrics
From Figure \ref{fig:violin_plot} and Table \ref{tab:aggregate_metrics}, we denote several meaningful trends. 
Using CSI as an overall proxy for skill of the FIM, we note that generally speaking the skill is correlated with a reduction of the stream orders of the processing units used for HAND.
In other words, the more you derive HAND on networks of unit drainage density and mosaic the resulting FIMs, the better those FIMs perform.
While this trend is evident for both sets of Manning's n values, the trend is slightly more significant for the higher value of 0.12.
Other trends related to this Figure include the general performance premium for 0.2\% events as opposed to lower 1\% events.
We also note how the higher Manning's n value enhances performance for both of these recurrence intervals across all models.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mapping Performance}
\label{ssec:mapping_performance}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We produced FIMs for the entire BLE domain within the 49 HUC8 areas across several states in the south central US. 
The forecasted FIMs using the discharges for the 1\% (100 year) and 0.2\% (500 year) recurrence flows directly from HEC-RAS were used to avoid noise and errors from hydrological processes.
We computed the statistics (CSI, POD, and FAR) for both 100 and 500 year events for Mannings N set to 0.06 and 0.12. 
The distribution of these statistics can be examined in Figure \ref{fig:violin_plot} as violin plots.
Each half of a violin plot represents the kernel density estimation (KDE) for a given model (FR, MS, GMS), given Manning's n value (0.06, 0.12), and given recurrence interval (1\%, 0.2\%), and performance metric (CSI, POD, FAR).
We also denote trend lines for each metric and Manning's n setting as well as their respective slope estimate and one-tailed p-value denoting the level of significance of the trend.

Aggregating the metrics in the method above treats each HUC8 as it's own unit and does little to consider the size differences of the HUCs. 
In an opposing aggregation method, we illustrate in Table \ref{tab:aggregate_metrics}  the CSI, POD, and FAR recomputed for the entire domain using the sum of all the TPs, FPs, and FNs. 
%
\begin{figure}[h!]
\centering
\includegraphics[scale=0.9]{figures/violin_plots.jpg}
\caption{Shows kernel density estimation of the distributions (sample size = 49) for 1\% (100 year) and 0.2\% (500 year) along with horizontal, dashed lines for the 25th, 50th, and 75th percentiles (in order from bottom to top).
The sub-figures separate the combination of three metrics (CSI, POD, and FAR) for two settings of Manning's n (0.06 and 0.12).
Trend lines for each metric - Mannings combination are shown (sample size = 294) along with associated slope and p-value of slope testing one-tailed significance.}
\label{fig:violin_plot}
\end{figure}
%
\begin{table}[h!]
\caption{Recomputed CSI, POD, and FAR using the primary metrics, TPs, FPs, and FNs, aggregated for BLE domain.
         Best value for across models is highlighted in bold.}
\label{tab:aggregate_metrics}
\centering
%\begin{tabular}{|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
\begin{tabular}{|c|c||c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Metric} & \multirow{2}{*}{Manning's n} & \multicolumn{2}{|c|}{FR} & \multicolumn{2}{|c|}{MS} & \multicolumn{2}{|c|}{GMS} \\
\cline{3-8}
  &  & 100yr & 500yr & 100yr & 500yr & 100yr & 500yr \\
\hline
\multirow{2}{*}{CSI} & 0.06 & 0.5576 & 0.5839 & 0.5717 & 0.5990 & \textbf{0.5796} & \textbf{0.6075} \\
\cline{2-8}
  & 0.12 & 0.5915 & 0.6149 & 0.6054 & 0.6288 & \textbf{0.6182} & \textbf{0.6435} \\
\hline
\multirow{2}{*}{POD} & 0.06 & 0.6354 & 0.6575 & 0.6524 & 0.6755 & \textbf{0.6633} & \textbf{0.6863} \\
\cline{2-8}
  & 0.12 & 0.7255 & 0.7446 & 0.7460 & 0.7648 & \textbf{0.7606} & \textbf{0.7810} \\
\hline
\multirow{2}{*}{FAR} & 0.06 & 0.1800 & 0.1609 & 0.1787 & \textbf{0.1589} & \textbf{0.1778} & \textbf{0.1589} \\
\cline{2-8}
  & 0.12 & 0.2379 & 0.2208 & 0.2374 & 0.2204 & \textbf{0.2324} & \textbf{0.2148} \\
\hline
\end{tabular}
\end{table}
%
% Interpretation of metrics
From Figure \ref{fig:violin_plot} and Table \ref{tab:aggregate_metrics}, we denote several meaningful trends. 
Using CSI as an overall proxy for skill of the FIM, we note that generally speaking the skill is correlated with a reduction of the stream orders of the processing units used for HAND.
In other words, the more we derive HAND on networks of unit drainage density and mosaic the resulting FIMs, the better those FIMs perform.
While this trend is evident for both sets of Manning's n values, the trend is slightly more significant for the higher value of 0.12.
Other trends related to this Figure include the general performance premium for 0.2\% events as opposed to lower 1\% events.
We also note how the higher Manning's n value enhances performance for both of these recurrence intervals across all models.
